Predicting the PCG PSeudo-Random Number Generator In Practice

### 2020-05-15 article/main.tex
section 2

@@

\todo{cette section doit décrire précisément PCG.}

\begin{quote}
  In addition, one of the generators, PCG-XSL-RR(described in Section 6.3.3), is
  explicitly designed to make any attempt at state reconstruction especially
  difficult, using xor folding tominimize the amount of information about
  internal state that leaks out.
\end{quote}

\begin{quote}
  Although these properties make it extremely challenging for an agent observing
  theexternal behavior of a program to guess the state of its random number
  generator,
\end{quote}

% import numpy
% print(numpy.random.default_rng())
% ---> Generator(PCG64)

\subsection{LCG algorithm}

The PCG algorithm we studied is based on the Linear Congruential Generators algorithm. The LCG algorithm is simply based on a congruential multiplication. This type of generator is based on four values: an internal state $S_{i}$, a multiplier $A$, an increment $C$ and the size $M$. Generally, to initialize a generator of a given size, a seed value is provided as the first internal state, and a default multiplier and increment are used. We will study the case where $M$ and $A$ are known and where $M$ is a power of 2.\\

Calling the generator results in calculating the next internal state $S_{i+1} = A \times S_i + C \mod{M}$ and outputting $S_{i+1}$ as data.

This type of generator is trivially predictable because, the internal state is given to us at each output, and in the event that $C$ is unknown, we can still deduce it with two consecutive outputs: $C = S_{i+1} - A \times S_i$.\\

A better variant of the standard LCG is the Truncated LCG, where the output consists only of the higher bits of the internal state $S_i$. This way the entire internal state isn't accessible. Returning the lower bits is not interesting because as we can understand easily and will prove later, lower bits of a state only depend on the lower bits of the previous state.

\subsection{PCG algorithm}

The PCG algorithm we studied is based on the LCG algorithm. To start, here are a few definitions:
\begin{itemize}
    \item let $k$ be the size of our generator, in this case $k = 128$ bits;
    \item let $M$ be the modulus with $M = 2^k$;
    \item let $S_i$ be the internal state at the $i$-th iteration and $S$ be the list of $S_i$;
    \item let $X_i$ be the output at the $i$-th iteration of the generator and $X$ be the list of $X_i$;
    \item let $lowS_i$ be the $k/2$ least significant bits of $S_i$;
    \item let $upS_i$ be the $k/2$ most significant bits of $S_i$;
    \item let $rot_i$ be the 6 most significant bits of $S_i$;
    \item let $nbiter$ be the number of consecutive outputs for prediction purposes.
\end{itemize}

The internal state can be redefined as:
\begin{equation}
    S_i = 2^{k/2} \times upS_i + lowS_i 
\end{equation}

The 6 most significant bits, $rot_i$, of the internal state are used as a value to determine the rotation in the PCG algorithm. Let \texttt{rotate} be the function that performs said rotation (Figure \ref{pcg128out}).\\
The rotation is a simple exchange of bits, where a $k/2$ bit value is split into 2 parts: one of $rot_i$ bits and the other in $(k/2) - rot_i$ bits. The two parts are then swapped, and the new value is returned.

This generator is initialised with a given $k$ bit internal state $S_0$ (of size 128 bits in the case of PCG128), and optionally with an $k$ bit increment $C$. If no increment is given, the algorithm will use a known default increment. For each output of the generator, the algorithm will perform the following calculations" (Figure \ref{pcg128out}):\\
\begin{align}
    X_{i} &= \mathtt{rotate}(upS_i\ \oplus\ lowS_i)\\
    S_{i+1} &= A \times S_{i} + C \mod{2^k}
\end{align}
In the case of PCG128, $A$ is a default increment of size 125 bits prime with $2^k$.

\begin{figure}[h!]
    \centering
    \includegraphics[width=0.75\linewidth]{pictures/PCG128.png}
    \caption{PCG128 : Output process}
    \label{pcg128out}
\end{figure}

Given the nature of the generator, the following values are known at every step:
\begin{itemize}
    \item the size of the generator $k$;
    \item the list of outputs $X$;
    \item the multiplier $A$;
    \item and eventually $C$, if the default increment was used. 
\end{itemize}

To break this generator, we had to differentiate between cases when the default $C$ was used or not. It was also interesting to note what was the minimum number of $X_i$ needed to reliably attack it.

@@

\todo{Comme on utilise à plusieurs reprise l'idée de se ramener à un LCG
  tronqué, autant décrire la technique à base de réseaux un bon coup.}
Pour casser PCG, nous avons utilisé la méthode présentée par Alan M. Frieze, Johan Hastad, Ravi Kannan, Jeffray C. Lagarias et Adi Shamir \cite{Frieze}. Plaçons nous dans le cas simple du Lehmer's generator\cite{Lehmer}. It's a Linear Congruential Generator, a known multiplier $A$, a known modulus $2^128$ and no increment. The outputs $X$ are the 64 upper-bits of $S$. For $n \in \mathbb{N}$, the $n$ firsts internal states form a lattice we will name $\mathcal{L}(A,2^{128},n)$. 
\todo{rajouter ref Lehmer's generator}

@@

\section{lattice geometric progression}

\subsection{The lattice}

@@

\subsection{A first way to solve the CVP : the Babai's rounding}

@@

By definition, \(r\) is the closest integer vector to \(TG^{-1}\). But as \(U\) is an element of the lattice, \(UG^{-1}\) is an integer vector. Thus \(r-TG^{-1}\) is smaller than \(UG^{-1}-TG^{-1}\).

@@

If we can prove that the left side of the inequality is smaller than the SVP of G, then wa have proved that \(rG\) is indeed the vector we were searching for.

>>>>

\DeclareMathSymbol{:}{\mathord}{operators}{"3A}

@@

This section describes the \textsf{PCG64} non-cryptographic pseudo-random number
generator (more precisely \textsf{PCG-XSL-RR} in the designer's terminology). It
has an internal state of 128-bit, which operate as a linear congruential
generator modulo~$2^{128}$. More precisely:
\[
  S_{i+1} = a S_i + c \bmod 2^{128},
\]
Where $a = 47026247687942121848144207491837523525$ is a fixed 126-bit
constant. The first initial state $S_0$ is the seed of the generator. The
increment $c$ may be specified by the user of the PRNG to produce different
output streams with the same seed (just as the IV acts in a stream cipher). If
no value of $c$ is specified, then a default increment is provided:
$c = 117397592171526113268558934119004209487$.

Each time the PRNG is clocked, 64 output bits are extracted from the internal
state using a non-linear function that makes use of data-dependent rotations, in
a way reminiscent of the \textsf{RC5} block cipher~\cite{Rivest94}. The six most
significant bits of the internal state encode a number $0 \leq r < 64$. The two
64-bit halves of the internal state are XORed together, and this 64-bit result
is rotated right by $r$ positions. The successive 64-bit outputs of the
generator are $X_0, X_1, \dots$ where:
\[
  X_i =(S_i[0:64] \oplus S_i[64:128]) \ggg S_i[122:128].
\]

Figure~\ref{pcg128out} summarizes the process. The overall design strategy is
similar to that of a filtered LFSR: the successive states of a weak internal
generator with a strong algebraic structure are ``filtered'' by a non-linear
function.

Updating the internal state requires a $128 \times 128 \rightarrow 128$
multiplication operation. In fact, this can be done with two
$64 \times 64 \rightarrow 64$ multiplications, one
$64 \times 64 \rightarrow 128$ multiplication and two 64-bit additions. High-end
desktop CPUs all implement these operations in hardware, so the generator is
relatively fast on these platforms.

\begin{figure}
  \begin{center}
\begin{tikzpicture}[scale=0.66]
%  \draw[red, use as bounding box] (-1.25, -6.75) rectangle (12.8, 2.25);

  % S_i

    % bordures
    \draw[very thick]  (0, 0) rectangle (12.8, 1);
    \draw  (0.6, 0) rectangle +(0, 1);
    \draw[very thick]  (6.4, 0) rectangle +(0, 1);
    \foreach \i in {0, 1, ..., 128} \draw (\i/10, 0) -- +(0, 0.1);

    % déco autour
    \node[draw] at (6.4, 3) (update) {$\times a + c \bmod 2^{128}$};
    \draw[thick,->] (12.8, 0.5) -- (13.8, 0.5) node[below right] {$S_i$} -- (13.8, 3) -- (update);
    \draw (13.7, 2.5) -- +(0.2, -0.2);
    \path (13.9, 2.5) node[anchor=west] {128};

    \draw[thick,->] (update) -- (-2, 3) -- (-2, 0.5) node[below left] {$S_{i+1}$} -- (0, 0.5);
    \draw (-2.1, 2.5) -- +(0.2, -0.2);
    \path (-1.9, 2.5) node[anchor=west] {128};

    \node[font=\footnotesize,anchor=east] at (12.9, -0.33) {0};
    \node[font=\footnotesize,anchor=east] at (6.5, -0.33) {64};
    \node[font=\footnotesize,anchor=west] at (0.5, -0.33) {122};
    \node[font=\footnotesize] at (-0.25, -0.33) {128};

    \draw (6.4, -2) node (x) {$\bigoplus$};

    \draw[->] (3.2, -0.1) |- (x);
    \draw (3.1, -0.9) -- +(0.2, -0.2);
    \path (3.3, -1) node[anchor=west] {64};

    \draw[->] (9.6, -0.1) |- (x);
    \draw (9.5, -0.9) -- +(0.2, -0.2);
    \path (9.7, -1) node[anchor=west] {64};

    \node[minimum width=1.8cm] at (6.4, -4) (r) {$\ggg$};
    \draw (5.8, -3.6) rectangle +(1.2, -0.8);
    \draw[fill=black] (5.8, -4.1) -- (5.95, -4) -- (5.8, -3.9) -- cycle;

    \draw (0.3, 1.1) -- (0.3, 1.5) -- (-1, 1.5) -- (-1, -4) -- (5.8, -4);
    \draw (-1.1, -2.9) -- +(0.2, -0.2);
    \path (-0.9, -3) node[anchor=west] {6};

    \draw[->] (x) -- +(0, -1.5);
    \node at (6.4, -6.25) (xi) {$X_i$};
    \draw[->] (6.4, -4.4) -- (xi);

    \draw (6.3, -4.9) -- +(0.2, -0.2);
    \path (6.5, -5) node[anchor=west] {64};
  \end{tikzpicture}
\end{center}
\caption{\textsf{PCG64}: Internal state update and output process.}
\label{pcg128out}
\end{figure}

@@

\todo{à fusionnier avec la section suivante!}
Pour casser PCG, nous avons utilisé la méthode présentée par Alan M. Frieze, Johan Hastad, Ravi Kannan, Jeffray C. Lagarias et Adi Shamir \cite{Frieze}. Plaçons nous dans le cas simple du Lehmer's generator\cite{Lehmer}. It's a Linear Congruential Generator, a known multiplier $A$, a known modulus $2^128$ and no increment. The outputs $X$ are the 64 upper-bits of $S$. For $n \in \mathbb{N}$, the $n$ firsts internal states form a lattice we will name $\mathcal{L}(A,2^{128},n)$. 
\todo{rajouter ref Lehmer's generator}

@@

\section{Reconstructing Truncated Geometric Sequences}
\todo{à fusionnier avec la section précédente!}
\subsection{The lattice}

@@

\subsection{A first way to solve the CVP : Babai's Round-off Algorithm}

@@

By definition, \(r\) is the closest integer vector to \(TG^{-1}\). But as \(U\) is an element of the lattice, \(UG^{-1}\) is an integer vector. Thus \(r-TG^{-1}\) is shorter than \(UG^{-1}-TG^{-1}\).

@@

If we can prove that the left side \todo{right side?} of the inequality is smaller than the SVP of G, then wa have proved that \(rG\) is indeed the vector we were searching for.
